---
title: "spatial join presung"
author: "Rachel Marconi"
date: "2023-11-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(sf)
library(tidyverse)
library(tigris)
library(janitor)
library(dplyr)

```


```{r get-census-tracts}
tract_set = tracts(
    state = "NC",
    county = "Mecklenburg"
    ) %>% 
  clean_names() %>% 
  select(-statefp, -countyfp, -aland, -awater, -funcstat, -mtfcc)

```


```{r more-school-data}
more_school_data <- read_csv('edp_school_data.csv', col_types = cols(.default="?", "ncessch" = col_character())) %>% 
  mutate(student_teacher_ratio = round(enrollment/teachers_fte,1)) %>% 
  mutate(percent_frl_eligible = free_or_reduced_price_lunch / enrollment) %>% 
  select(-year, -state_location, -charter)

race_school_data <- read_csv('edp_school_race.csv', col_types = cols(.default="?", "ncessch" = col_character())) %>% 
  pivot_wider(names_from = race, values_from = enrollment) %>% 
  clean_names() %>% 
  mutate(share_white = round(white/total*100), 
         share_black = round(black/total*100),
         share_hisp = round(hispanic/total*100), 
         share_asian = round(asian/total*100), 
         share_other = round((american_indian_or_alaska_native + native_hawaiian_or_other_pacific_islander + two_or_more_races)/total*100)) %>% #unknown all zero
  select(-year, -school_name, -state_location, -total) 

school_specific_data <- more_school_data %>% 
  left_join(race_school_data, by=c("ncessch" = "ncessch")) %>% 
  st_as_sf(coords = c("longitude", "latitude"),
            crs = st_crs(tract_set))
```



###### SCHOOL DATA INTERLUDES

###### interlude to find longest school name for label widths
```{r}
school_names <- more_school_data %>% 
  mutate(length = nchar(school_name))
# it's South Academy of International Languages
```

###### interlude to make marker layer geojson
###### just basic data
```{r markers}
school_markers <- school_specific_data %>% 
  select(ncessch, school_name, school_level, geometry)

# this is pulled in to place markers on the map in map.js
st_write(school_markers, "out/school_markers.geojson", append=FALSE)
```

###### interlude to save out school race data alone for in dot charts
```{r}
school_race_alone <- race_school_data %>% 
  select(ncessch, contains("share"))

#this is converted to json and pulled in as school race/eth source of truth for dot charts in the multiple race/eth components
write_csv(school_race_alone, "out/school_race_alone.csv")
```


###### public housing data and market share from Tina
```{r get-market-share-data}
tract_housing_market_share <- read_csv("Mecklenburg_newVars.csv") %>% 
  clean_names() %>% 
  mutate(li_share = ifelse(is.na(li_housing_share), 0, li_housing_share)) %>% 
  mutate(percent_renter_occupied = round(percent_renter_occupied*100),
         percent_owner_occupied = round(percent_owner_occupied*100)) %>% 
  select(geoid, market_share_total,percent_renter_occupied, percent_owner_occupied, li_share, tenure_total)

```


###### combines school data to their tract data, unique schools
```{r join-to-tracts}
joined_school_data <- st_join(school_specific_data, tract_set) %>% 
  mutate(scl_pct_nonwhite = round((enrollment - white)/enrollment*100,1),
         share_frl_eligible = round(percent_frl_eligible*100,1))
  

```

###### tract data alone, unique tracts, manipulate into tool ready units, combos and rounding
```{r get-tract-data}
tract_data <- read_csv('CMSD_tract_2021.csv') %>% 
  clean_names() %>% 
  mutate(tract_pct_nonwhite = round( (1 - percent_white_nonhisp)*100,1)) %>% 
  mutate(test_sfh_pct = round(x1_unit_detached_home / housing_units * 100,0)) %>% 
  select(-percent_renter_occupied, -percent_owner_occupied) %>% 
  left_join(tract_housing_market_share, by=c("geoid"="geoid")) %>% 
   mutate(school_age_kids_in_tract = rowSums(select(., contains("under5") , contains("5to9") , contains("10to14") , contains("15to17"))))

tract_indicators <- tract_data %>% 
    mutate(percent_owner_occupied = round(percent_owner_occupied *100,0),
         percent_renter_occupied = round(percent_renter_occupied *100,0),
         market_share = round(market_share_total *100,0)) %>% 
  select(geoid, market_share,school_age_kids_in_tract,
         test_sfh_pct, median_gross_rent, median_owner_cost_mortgage, percent_owner_occupied, percent_renter_occupied) %>% 
  write_csv("out/tract_indicators_only.csv")
#this csv is converted to json and used in currentTract.js store for selected tract data in the tool

county_indicators <- tract_indicators %>% 
  select(-geoid) %>% 
  colMeans(na.rm=TRUE)

county_school_aged_kids <- tract_indicators %>% 
  select(school_age_kids_in_tract) %>% 
  colSums()
  
county_indicators
#these are hardcoded into the Charlotte-specific components to reduce the number of files we're loading into the tool
```


###### interlude to save tract race and housing data to file
```{r}
tract_race <- tract_data %>% 
  mutate(all_kids_in_tract = rowSums(select(., contains("male_")))) %>% 
  mutate(share_white_kids = round(rowSums(select(., starts_with("white_")))*100/all_kids_in_tract),
         share_black_kids = round(rowSums(select(., starts_with("black_")))*100/all_kids_in_tract),
         share_hispanic_kids = round(rowSums(select(., starts_with("hisp_")))*100/all_kids_in_tract),
         share_asian_kids = round(rowSums(select(., starts_with("asian_")))*100/all_kids_in_tract),
         share_other_kids = round(rowSums(select(., starts_with("aian_"), starts_with("nhpi_"), starts_with("native_"), starts_with("other_")))*100/all_kids_in_tract)
      ) %>% 
  select(geoid, share_white_kids, share_black_kids, share_hispanic_kids, share_asian_kids, share_other_kids)

write_csv(tract_race, "out/raceeth_kids_in_tract.csv")
#this csv is encoded as json and used as the tract source of truth for dot charts in the multiple race/eth components


county_race <- tract_race %>% 
  select(-geoid) %>% 
  colMeans(na.rm=TRUE)





hud_data <- read_csv("hudPicture2023_383145.csv") %>% 
  clean_names() %>% 
  filter(trunc(code / 1000000) == 37119) %>% 
  select(code, subsidized_units_available)

tract_housing <- tract_data %>% 
  select(geoid, li_share, tenure_total) %>% #need to update with ph units
  left_join(hud_data, by = c("geoid" = "code")) %>% 
  mutate(li_share = round(li_share*100,1),
         ph_share = round(subsidized_units_available/tenure_total*100),
         ph_share = ifelse(ph_share == -Inf, 0, ph_share),
         ph_error = subsidized_units_available == -1,
         non_ph_units = 100-li_share-ph_share,
         non_ph_units = ifelse(non_ph_units > 100, 100, non_ph_units)) %>% 
  write_csv("out/housing_units_in_tract.csv")
#this csv is encoded as json and pulled into our tract housing gridboxes and data

county_housing <- tract_housing %>% 
  select(-geoid) %>% 
  colMeans() 

```



```{r join-tract-and-school-data}
joined_tract_school_data <- joined_school_data %>% 
  left_join(tract_data, by=c('tractce'='tract')) %>%
  st_drop_geometry()
 

#ultimately feeding the tool
all_indicators <- joined_tract_school_data %>% 
  select(geoid = geoid.x, ncessch, school_name, street_location, city_location,enrollment, school_level, school_type, school_age_kids_in_tract,tract_pct_nonwhite,market_share_total,
         scl_pct_nonwhite,
         share_frl_eligible, meps_poverty_pct,student_teacher_ratio, test_sfh_pct,
         median_gross_rent,median_owner_cost_mortgage,percent_renter_occupied,percent_owner_occupied) #%>% 
  #filter(school_level != "Missing/not reported")
  
write_csv(all_indicators, 'out/schools_indicators_only.csv')
# this is encoded as all_indicators.json and pulled into school.js for school data and to run the scatterplot
```


